Example visualization
====================

```{r opts,echo=FALSE}
opts_chunk$set(fig.height=5,fig.width=5)
```

```{r pkgs,message=FALSE}
library(data.table)  ## for fast fread()
library(ggplot2)  ## nice plots
theme_set(theme_bw())  ## cosmetic
library(KernSmooth)
```

Read data, organize, merge:

```{r getdat}
fn <- "ex1"
hdat <- fread(paste0(fn,".household.csv"))
idat <- fread(paste0(fn,".individual.csv"))
## should be unnecessary once flat files have column headings
setnames(idat, names(idat), c("hID", "sex", "age", "disStatus"))
setnames(hdat, names(hdat), c("hID", "x", "y", "income", "disRisk"))
## key both data sets on household ID and merge
setkey(hdat, hID)
setkey(idat, hID)
idat2 <- merge(hdat, idat)
```

R's plotting capabilities are not actually so great for dealing with
very large data sets (these count as "somewhat large") -- we really
need to summarize in some efficient way first, then plot ...

```{r plot1}
## plot(y~x,data=hdat,pch='.') ## not very informative!!
## with(hdat,plot(hexbin(x,y))) ## efficient but ugly nice except for edge
## effects ...
ggplot(hdat, aes(x, y)) + geom_hex() + 
     scale_x_continuous(expand = c(0, 0)) + 
    scale_y_continuous(expand = c(0, 0))
```

What we really want is to do hexagonal binning or kernel density estimation to get a reasonable smoothed version of the variable of interest (income, disease risk, etc.) so we can map it.  I think in this case everything's going to be pretty boring since we haven't built in very much interesting spatial variation yet ...

Hexagonal binning is nice; I need to work out 
[this solution](http://stackoverflow.com/questions/17688012/hexbin-apply-function-for-every-bin) and
then see if I can feed it `ggplot2::geom_hex` for prettier output ...


```{r hexbin1}
hex1 <- hexbin(idat2$x, idat2$y, IDs = TRUE)
```


Kernel smoothing:
```{r ksmooth1}
kk <- bkde2D(hdat[, list(x, y)], bandwidth = c(0.1, 0.1))
```

Base contour plot (nothing going on here!)

```{r contour1}
with(kk, contour(x1, x2, fhat))
```

```{r xbrks,results="hide"}
brks <- seq(min(hdat[,x]), max(hdat[,x]), length=51)
## and := to avoid an additional copy
## (can this be done in a single step?)
idat2[,xbin:=findInterval(x, brks)]
idat2[,ybin:=findInterval(y, brks)]
```

Aggregate with mean by bin; use `.N` to include number of elements per bin, and plot.
```{r bin}
idatsum <- idat2[, list(.N, d=mean(disStatus)), 
                  by=list(xbin,ybin)]
ggplot(idatsum,aes(xbin,ybin))+geom_tile(aes(fill=d))
```
Was going to plot average disease risk instead, but I think disease status wasn't implemented yet (no-one in this data set was infected)

```{r bin2}
hdatsum <- idat2[, j=list(x=x[1],y=y[1],
                          .N, dsum=sum(disStatus)), 
                  by=hID]
ggplot(idatsum,aes(xbin,ybin))+geom_tile(aes(fill=d))
```

Zoom in on a region:
```{r zoom}
hdatfoc <- hdatsum[0.5<x & x<0.6 & 0.5<y & y<0.6,]
ggplot(hdatfoc,aes(x,y))+geom_point(aes(size=N,colour=dsum))
```

## To do

### Pix

* hexbin plots? of income, disease risk, ...
* associations of covariates -- disease vs gender, age, income, etc.
* check out `bigvis` package?

### Data manipulation

* could compute household averages of disease, for faster plotting (should be easy with `data.table`)
                                                              
### Cosmetics

* get rid of `fread` progress bars
* fix `findInterval` edge cases?
